---
title: "Data Visualisation with ggplot2 Project Joseph Thiongo"
output: html_document

---

# AfterWork Data Science: Data Visualisation with Python Project

## 1. Defining the Question

### a) Specifying the Data Analysis Question

To provide Hilton management with strategy recommendations that will
lead to revenue growth.

### b) Defining the Metric for Success

The solutions to the following questions will help us answer our
research question:

-   When is the best time of year to book a hotel room?
-   When is the optimal length of stay in order to get the best daily
    rate?
-   How will you know if a hotel was likely to receive a
    disproportionately high number of special requests?

### c) Understanding the context

I am a Data Scientist working for Hilton International Hotels and need
to help the management decide on the strategies that will lead to
revenue growth.

I have a dataset containing information on the various chain of hotels,
including when customers made the booking, length of stay, the number of
adults, children, or babies, and the number of available parking spaces,
among other things.

Given this dataset, i will perform data exploration, data cleaning, and
analysis to come with appropriate recommendations. As a start, while
performing data analysis, i will ask the following questions from the
given dataset. Making sure to have a visualization for each question. ●
How many bookings were canceled? ● What was the booking ratio between
resort hotels and city hotels? ● What was the percentage of booking for
each year? ● Which was the busiest month for hotels? ● Most guests came
from which country? ● How long do most people stay in hotels? ● Which
was the most booked accommodation type (Single, Couple, Family)?

### d) Recording the Experimental Design

The steps to be taken include:

-   Load dataset and preview its summarized information to get a feel of
    what i will be working with.
-   Carry out data cleaning.
-   Carry out data analysis.
-   Interpret results.
-   Provide recommendations based on results of analysis.
-   Challenge your solution.

### e) Data Relevance

The dataset contains information regarding customer bookings which will
aid our research analysis

## 2. Reading the Data

```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
library("tidyverse")
```

```{r}
# Load the data below
# --- 
# Dataset url = 
# --- 
# YOUR CODE GOES BELOW
# 

source_df <- read_csv("hotel_bookings.csv")
head(source_df)

```

```{r}
# Checking the first 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 

head(source_df,5)
```

```{r}
# Checking the last 5 rows of data
# ---
# YOUR CODE GOES BELOW
# 
tail(source_df,5)
```

```{r}
# Sample 10 rows of data
# ---
# YOUR CODE GOES BELOW
# 
sample_n(source_df,10)
```

```{r}
# Checking number of rows and columns
# ---
# YOUR CODE GOES BELOW
#  
dim(source_df)
```

```{r}
# Checking datatypes
# ---
# YOUR CODE GOES BELOW
# 

str(source_df)
```

```{r}
glimpse(source_df)
```

Record your general observations below:

-   Observation 1 - Dataset has 119,390 observations
-   Observation 2 - Dataset has 32 feature columns. Some columns have
    'NULL' entries

## 3. External Data Source Validation

The data is originally from the article Hotel Booking Demand Datasets,
by Nuno Antonio, Ana Almeida, and Luis Nunes for Data in Brief, Volume
22, February 2019.

## 4. Data Preparation

### Performing Data Cleaning

```{r}
# Checking datatypes and missing entries of all the variables
# ---
# YOUR CODE GOES BELOW
# 
glimpse(source_df)

```

We observe the following from our dataset:

Observation 1 - Column [is_canceled] ..uses 0,1 to represent YES, NO

Observation 2 - Column [company]... has so many 'NULL' entries, might need to drop it

```{r}
# Checking how many duplicate rows are there in the data
# ---
# YOUR CODE GOES BELOW
# 

#dim(source_df)
#unique(source_df)
src1 <- source_df[!duplicated(source_df), ]
dim(src1)

```

We choose to keep the duplicates because we don't have a unique
identifier to actually proof that we have duplicates.

```{r}
# Checking if any of the columns are all null
# ---
# YOUR CODE GOES BELOW
# 

colSums(is.na(source_df))
```

We observe the following from our dataset:

Observation 1 - Out of 119,390 observations we have 87,396 unique ones.
However, there is no unique identifier to confirm that there is no data
duplication. Only column children as 4 missing values

```{r}
# Checking if any of the rows are all null
# ---
# YOUR CODE GOES BELOW
# 

#rowSums(is.na(source_df))
is.null(source_df)
```

We observe the following from our dataset:

Observation 1 - None of the rows are all NULL Observation 2 - None of the columns are all null

```{r}
install.packages("reshape2", repos = "http://cran.us.r-project.org")
#loading the package
library(reshape2)
```

```{r}
# Checking the correlation of the features through the use of 
# visualizations the correlation using heatmap
# ---
# YOUR CODE GOES BELOW
# 


# Select the desired columns
src2_df <- source_df %>%
  na.omit() %>%
  select(is_canceled, lead_time,	arrival_date_year, arrival_date_week_number, arrival_date_day_of_month, stays_in_weekend_nights, stays_in_week_nights, adults, children,babies, is_repeated_guest,	previous_cancellations,	previous_bookings_not_canceled,booking_changes,days_in_waiting_list,adr,required_car_parking_spaces,total_of_special_requests)
#src2_df


# Create a correlation matrix
corr_matrix <- cor(src2_df, method="s")
head(corr_matrix)

# Create a table with long form
corr_df <- melt(corr_matrix)
head(corr_df)


# Plot the heatmap
ggplot(corr_df, aes(Var1, Var2, fill = value)) + 
  geom_tile(color = "black") + 
  geom_text(
    aes(label = round(value, 2)), 
    color = "white"
  ) +
  coord_fixed() + 
  labs(
    fill="Pearson Correlation"
  ) +
  scale_fill_gradient2(
    low = "blue", 
    high = "red",
    mid = "white", 
    limit = c(-1,1)
  ) + 
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

We observe the following from our dataset:

Observation 1- There is some correlation between feature 'previous guest
not cancelled' and 'is repeated guest' Observation 2- The other features
display low +ve or -ve correlation

```{r}
# Dropping company column because it has alot of missing values 
# and we won't need to answer any of our questions
# ---
# YOUR CODE GOES BELOW
# 

source1_df = select(source_df, -c("company"))
#dim(source1_df)
head(source1_df)
```

From the data variable description we see that the Distribution Channel
category that tells us about Booking distribution.

The term "TA" means "Travel Agents" The term "TO" means "Tour Operators"
This allows us to fill the missing values in the agents column with TO

```{r}
# We replace the mising values i.e. for TO
# ---
# YOUR GOES BELOW
#

source1_df$agent[source1_df$agent == 'NULL'] <-240 
#unique(source1_df$agent)
```

```{r}
# We drop rows where there is no adult, baby and child as 
# these records won't help us.
# ---
# YOUR GOES BELOW
#

abc1_filter <- which(source1_df$adults ==0 & source1_df$children ==0 & source1_df$babies ==0)
source1_df <- source1_df[-abc1_filter,]
dim(source1_df)

```

```{r}
# We replace missing children values with rounded mean value
# ---
# Hint i.e. use round()
# ---
# YOUR GOES BELOW
#

source1_df$children[is.na(source1_df$children)] <- round(mean(source1_df$children, na.rm = TRUE))
#unique(source1_df$children)
```

```{r}
# Checking for missing values in the dataframe
# ---
# YOUR GOES BELOW
#
colSums(is.na(source1_df))
```

```{r}
# Converting the datatypes of the following columns from float to integer
# i.e. children, company, agent
# ---
# YOUR GOES BELOW
# 

#str(source1_df)
source1_df$children <- as.integer(source1_df$children)
source1_df$agent <- as.integer(source1_df$agent)
str(source1_df)
```

## 5. Solution Implementation

### 5.a) Questions

```{r}
# 1. How many bookings were cancelled?
# ---
# Visualisation: Barplot


title1 <- paste(length(which(source1_df$is_canceled ==1)), ":bookings were cancelled")
             
# Create the summary table
source1_df %>% 
  group_by(arrival_date_month) %>%
  summarize(is_canceled = length(which(source1_df$is_canceled ==1)))

# Plotting the chart
ggplot(source1_df, aes(x = arrival_date_month, y = is_canceled)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = title1, 
    x = "Month", 
    y = "Cancellations"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
# 2. What was the booking ratio between resort hotel and city hotel?
# ---
# Barplot of booking ratio between resort hotel and city hotel

#install.packages("sqldf")
library(sqldf)

sub1_df <- sqldf("SELECT hotel, count(hotel) as bookings  FROM source1_df GROUP BY hotel")
sub1_df


# Plotting the chart
ggplot(sub1_df, aes(x = hotel, y = bookings)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  geom_text(aes(label=bookings), position=position_dodge(width=0.9), vjust=-0.25) +
  labs(
    title = "Bookings ratio - City Hotel vs Resort Hotel", 
    x = "Hotel", 
    y = "Bookings"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
# 3. What was the percentage of booking for each year?
# ---
# 

#sum1 <- length(which(source1_df$is_canceled ==1))
sum1 <- length(source1_df$is_canceled)

# Stacked + percent
ggplot(source1_df, aes(fill=hotel, y=sum1, x=arrival_date_year)) + 
    geom_bar(position="fill", stat="identity") +
    scale_y_continuous(labels = scales::percent,
                     breaks = scales::pretty_breaks(n = 8)) +
    labs(
    title = "Percentage bookings for each year", 
    x = "Year", 
    y = "Booking percentage"
  ) 

```

```{r}
# 4. Which were the most busiest months for hotels?
# ---
# 



subs2_df <- sqldf("SELECT count(reservation_status) as checkouts, arrival_date_month as month FROM source1_df WHERE reservation_status == 'Check-Out' GROUP BY arrival_date_month")

#subs2_df

# Plotting the chart
ggplot(subs2_df, aes(x = month, y = checkouts)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Busiest months for hotels", 
    x = "Month", 
    y = "Checkouts count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

```{r}
# 5. From which top 3 countries did most guests come from?
# --- 
# YOUR GOES BELOW
#

subs3_df <- sqldf("SELECT country, count(country) as cnt1 FROM source1_df GROUP BY country ORDER BY cnt1 DESC LIMIT 5")
subs3_df

# Plotting and styling
ggplot(data=subs3_df, aes(x="", y=cnt1, fill=country)) +
  geom_bar(
    stat="identity", width=0.5, color = "white"
  ) +
  coord_polar("y") + 
  geom_text(
    aes(label = paste0(round(cnt1 / sum(cnt1) * 100, 1), "%")), position = position_stack(vjust = 0.5)
  ) +
  scale_fill_brewer(palette="Set1") +   
  labs(
    fill = "Country", 
    title = "Which countries do most guests come from?"
  ) +
  theme_void()


```

```{r}
# 6.a) How long do most people stay in hotels?
#   b) By city and resort? Separate the data by hotel
# ---
# 

subs4_df <- sqldf("SELECT (stays_in_weekend_nights + stays_in_week_nights) as totalstays FROM source1_df WHERE hotel == 'Resort Hotel'")
#subs4_df

subs5_df <- sqldf("SELECT (stays_in_weekend_nights + stays_in_week_nights) as totalstays FROM source1_df WHERE hotel == 'City Hotel'")
#subs5_df


ggplot(subs4_df, aes(x = totalstays)) +
  geom_histogram(
    bins=30, fill = "orange", color = "white"
  ) + 
  scale_x_continuous(breaks = seq(0,60,5))+
  labs(
    title = "Distribution of total stays in Resort hotel", 
    x = "Total stays", 
    y ="Frequency"
  ) +
  theme(
    plot.title = element_text(color = "#0099f9", size = 15, face = "bold", hjust = 0.5), 
  )



ggplot(subs5_df, aes(x = totalstays)) +
  geom_histogram(
    bins=30, fill = "green", color = "white"
  ) + 
  scale_x_continuous(breaks = seq(0,60,5))+
  labs(
    title = "Distribution of total stays in City hotel", 
    x = "Total stays", 
    y ="Frequency"
  ) +
  theme(
    plot.title = element_text(color = "#0099f9", size = 15, face = "bold", hjust = 0.5), 
  )


```

```{r}
# 7. Which was the most booked accommodation type (Single, Couple, Family)?
# ---
#

ggplot(source1_df, aes(x = reserved_room_type)) +
  geom_histogram(
    bins=20, fill = "red", color = "white", stat="count"
  ) + 
  labs(
    title = "Which was the most booked accommodation type?", 
    x = "Reserved Room Type", 
    y ="Frequency"
  ) +
  theme(
    plot.title = element_text(color = "#0099f9", size = 15, face = "bold", hjust = 0.5), 
  )

```

### 5.b) Recommendations

From the above analysis, below are our recommendations:

-   Bulk of bookings come from 3 countries [PRT, GBR, FRA]. Greater
    investment in marketing and promotion in this 3 countries will boost
    revenue
-   Most popular reserved room types are A and D. Room Types B,C,F,G,H,L
    have such low demand that management should consid converting them
    to room types A,D or E
-   The busiest months are July and August. Hotels staffing and
    resourcing during these months should be optimized to cater for
    surges in demand with no reduction in service quality due to high
    customer traffic
-   City Hotels have significantly more bookings than resort hotels.
    Core business functions should be focused more on the City Hotel
    side of business than the Resorts

## 6. Challenging your Solution

In this step, we review our solution and implement approaches that could
potentially provide a better outcome. In our case, we could propose the
following question that wasn't answered in our solution because it
couldn't have greatly contributed to our recommendation.

```{r}
# When should hotels provide special offers?
# ---
# YOUR GOES BELOW
# 

# Plotting the chart
ggplot(subs2_df, aes(x = month, y = checkouts)) +
  geom_col(
    stat="identity", width=0.5, fill = "#0099f9"
  ) +
  labs(
    title = "Busiest months for hotels", 
    x = "Month", 
    y = "Checkouts count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size=16, family="Times New Roman"), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

-   Our observations:August and July months have the highest bookings
    traffic They are ideally suited for any promotional and special
    offers activity

-   How does this observation tie to our solution?... July and August
    are the busiest months. Have already recommended that activities
    geared toward capturing all demand and attracting new customers
    should be enhanced during these months to boost sales

## 7. Follow up questions

During this step, you rethink and propose other ways that you can
improve your solution.

a). Did we have the right data? ...Yes, we had the right data 
b). Do we need other data to answer our question?... No, we were able to advise
management on best methods to boost sales/revenues 
c). Did we have the right question?...Yes, the research questions helped in providing
recommendations to management






